---
title: "Olist E-Commerce Customer Segmentation"
author: "Meera Sanjeevirao"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
# This chunk sets up the document and loads all necessary packages.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
library(factoextra)
library(stats)
library(knitr)
```

## 1. Introduction and Data Preparation

This report details the customer segmentation for the Olist e-commerce platform. We use K-Means clustering in R to identify distinct customer groups based on their RFM (Recency, Frequency, Monetary) behavior.

The foundation of this analysis is the `rfm_data.csv` file, which was generated by the `calculate_rfm.sql` script. This script transformed raw transactional data by joining customer, order, and payment tables to compute the core RFM metrics for each unique customer:

- Recency (R): The number of days between the customer's most recent order and the latest transaction date in the dataset.

- Frequency (F): The total count of completed orders for each customer.

- Monetary (M): The total sum of all payments made by each customer.

Once this aggregated RFM data is loaded into R, further preparation is needed. Because RFM distributions are often skewed, we apply a log transformation. The data is then scaled to give each metric equal importance during the clustering process, as the K-Means algorithm is sensitive to variations in feature scales.

```{r data_prep}
# Load the dataset created from the SQL query
rfm_data <- read.csv("../data/processed/rfm_data.csv")

# Log-transform and scale the data for clustering
rfm_prepared <- rfm_data %>%
  # Add 1 to avoid log(0) issues for customers with 0 recency.
  mutate(
    log_recency = log(recency + 1),
    log_frequency = log(frequency + 1),
    log_monetary = log(monetary + 1)
  )

rfm_scaled <- scale(rfm_prepared[, c("log_recency", "log_frequency", "log_monetary")])
```

## 2. K-Means Clustering

### Determining Optimal Clusters
To find the optimal number of segments, we use the Elbow Method. This technique calculates the Total Within-Cluster Sum of Squares (WSS) for different cluster counts (k). We are looking for the "elbow" on the plot—the point where the WSS begins to decrease at a much slower rate. This point indicates the most appropriate number of clusters.

The plot below shows a distinct elbow at k=4, making 4 clusters a reasonable choice for this analysis.

```{r optimal_clusters}
# Calculate the Within-Cluster Sum of Squares (WSS) manually for a range of k values to create our own Elbow Plot.

# Function to compute total WSS for a given k
calculate_wss <- function(k, data) {
  kmeans(data, k, nstart = 25)$tot.withinss
}

# Set the range of clusters to test (e.g., 1 to 10)
k_values <- 1:10

# Calculate WSS for each k value
# This loop is memory efficient as it processes one k at a time
wss_values <- sapply(k_values, calculate_wss, data = rfm_scaled)

# Create a data frame for plotting
elbow_data <- data.frame(
  clusters = k_values,
  wss = wss_values
)

# Plot the elbow curve using ggplot2
ggplot(elbow_data, aes(x = clusters, y = wss)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  labs(
    title = "Elbow Method for Optimal Number of Clusters",
    subtitle = "Calculated Manually",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster Sum of Squares (WSS)"
  ) +
  scale_x_continuous(breaks = k_values) +
  theme_minimal()
```

### Running the Algorithm
We now run the K-Means algorithm with 4 centers and add the cluster assignments back to our data for analysis.

```{r kmeans}
set.seed(123) # for reproducibility
kmeans_result <- kmeans(rfm_scaled, centers = 4, nstart = 25)
rfm_prepared$cluster <- as.factor(kmeans_result$cluster)
```

## 3. Cluster Analysis & Interpretation

To understand what each cluster represents, we calculate the average RFM values for each group and assign a descriptive persona.

```{r cluster_summary}
# Calculate the mean RFM values for each cluster
cluster_summary <- rfm_prepared %>%
  group_by(cluster) %>%
  summarise(
    avg_recency = mean(recency),
    avg_frequency = mean(frequency),
    avg_monetary = mean(monetary),
    customer_count = n()
  ) %>%
  arrange(desc(avg_monetary))

# Assign meaningful persona names based on the characteristics
cluster_summary <- cluster_summary %>%
  mutate(persona = case_when(
    # Catches multi-purchase customers who are now less recent.
    avg_frequency > 1.5 & avg_monetary > 200   ~ "Loyal Customers",
    
    # Catches high-value, single-purchase customers who are now inactive.
    avg_recency > 250 & avg_monetary > 200   ~ "Hibernating High Spenders",
    
    # Catches recent, lower-value, single-purchase customers.
    avg_recency < 150                          ~ "New Customers",
    
    # Catches all other single-purchase, non-recent, low-value customers.
    TRUE                                       ~ "At-Risk Low Spenders"
  ))

# Display the summary table
kable(cluster_summary, caption = "Customer Segment Characteristics")
```

## 4. Visualizing the Segments

Visualizations help illustrate the distinct characteristics of our segments.

### RFM Segment Scatter Plot
This plot visualizes the distinct characteristics of our four segments. The x-axis (Recency) is on a log scale to better distribute the customers.

We can see a clear separation:
* **Loyal Customers** and **New Customers** are on the right side of the plot, indicating low recency (they've purchased recently). They are differentiated by their frequency and spending.
* **Hibernating High Spenders** and **At-Risk Low Spenders** are on the left (high recency). The key difference between them is their historical spending, which is represented by the size of the points.

```{r, fig.width=10, fig.height=7}
# Add the persona names back to the main data frame for plotting
rfm_prepared <- rfm_prepared %>%
  left_join(cluster_summary %>% select(cluster, persona), by = "cluster")

# Create the scatter plot
ggplot(rfm_prepared, aes(x = recency, y = frequency, color = persona, size = monetary)) +
  geom_point(alpha = 0.7) +
  scale_x_continuous(trans = 'log10') +
  labs(
    title = "Customer Segments by RFM",
    x = "Recency (Days, Log Scale)",
    y = "Frequency (Number of Orders)",
    color = "Persona",
    size = "Monetary Value"
  ) +
  theme_minimal()
```

## 5. Conclusion & Recommendations

The K-Means analysis successfully segmented Olist customers into four distinct and actionable groups. By replacing generic labels with personas that reflect the true nature of the clusters, we can develop more precise marketing strategies:

**Loyal Customers**: These are repeat buyers with high frequency and good monetary value. They are the backbone of the business, but since they haven't purchased as recently as new customers, the key is to prevent churn.

- Action: Target with loyalty programs, exclusive access to new products, and personalized "thank you" offers to reward their business and keep them engaged.

**Hibernating High Spenders**: This group represents customers who have spent a significant amount in the past but have not purchased in a long time. They are a high-value, high-risk segment with huge potential if won back.

- Action: Launch a targeted win-back campaign. Use personalized emails reminding them of past purchases and offer a compelling discount to encourage their return.

**New Customers**: This segment consists of recent, first-time buyers who have spent a modest amount. The primary goal is to nurture them into becoming loyal customers.

- Action: Encourage a second purchase through a welcome series, targeted follow-up promotions, and product recommendations based on their initial order.

**At-Risk Low Spenders**: This is the largest group, containing customers who made a single, low-value purchase a long time ago and have not returned.

- Action: This group has a low probability of converting. It is not cost-effective to spend significant marketing resources here. A low-touch, automated email campaign is sufficient.

This data-driven segmentation provides a strong foundation for developing targeted marketing strategies to improve customer retention and maximize lifetime value.